Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2019-10-25 write-up/fda-recurrent.tex
Updated paper

\def\S{\mathcal{S}}
\def\indep{\mathrel{\rlap{$\perp$}\kern1.6pt\mathord{\perp}}}
\def\R{\mathcal{R}}
\def\H{\mathcal{H}}
\def\E{\mathbb{E}}
\def\Y{{\bf Y}}
\def\Cov{\text{Cov}}
\def\one{{\bf 1}}
\def\diag{\text{diag}}
\def\given{\, | \,}
\def\Given{\, \big | \,}
\def\Nat{\mathbb{N}}
\def\Real{\mathbb{R}}
\def\bft{{\bf t}}
\def\bfx{{\bf x}}
\def\bfp{{\bf p}}
\def\bfT{{\bf T}}
\def\dotminussym#1#2{%
  \setbox0=\hbox{$\m@th#1-$}%
  \kern.5\wd0%
  \hbox to 0pt{\hss\hbox{$\m@th#1-$}\hss}%
  \raise.6\ht0\hbox to 0pt{\hss$\m@th#1.$\hss}%
  \kern.5\wd0}
\newcommand{\dotminus}{\mathbin{\mathpalette\dotminussym{}}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

Let~$H_{i,t}^{NX} = H_{i,t}^{N} \otimes H_{i,t}^{X}$ be the $\sigma$-field generated by all past values~$(N_i (s), x_i (s))_{0 \leq s \leq t}$. The instantaneous risk of an event at time~$t$ is assumed to depend on the health process, time-in-study, and the event history through a fully parametric conditional hazard function:
\begin{equation}
\label{eq:hazard}
h_i \left( t \Given H_{i,t}^{NX} ; \theta \right) =
\lim_{\delta \to 0} \delta^{-1} \pr \left( N_i(t+\delta) - N_i(t) > 0
  \given H_{i,t}^{NX}; \theta \right),
\end{equation}
where~$\theta$ is the parameter vector. Equation~\eqref{eq:hazard}
states the hazard may depend on past values of the event history and
covariates.  For high frequency physiological data, we assume that
current risk is log-additive and depends on a linear functional of the
health process over some recent window of time and some pre-specified
features of the counting process; that is,
\begin{equation}
\label{eq:hazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_i (s) \beta(s) ds  \right)
\end{equation}
where~$\Delta$ is an unknown window-length, and $g_t( H_{i,t}^N ) \in
\mathbb{R}^p$ is a $p$-length feature vector summarizing the
event-history and time-in-study information. The final
term~$\int_{t-\Delta}^t x_i(s) \beta(s) ds$ reflects the unknown
linear functional form of the impact of the time-varying covariate on
current risk.

An alternative to~\eqref{eq:hazardlinear} would be to construct
features from the sensor data history~$f_t ( H_{i,t}^{X}) \in
\mathbb{R}^q$ and incorporated these features in the place of the
final term. Our current approach builds linear features of~$H_{i,t}^X$
directly from the integrated history, avoiding the feature
construction problem -- a highly nontrivial issue for high frequency
time-series data.  The main caveat is the additional
parameter~$\Delta$; however, as long as the estimated~$\hat \Delta$
exceeds~$\Delta$, then resulting estimation is unbiased albeit at a
loss of efficiency.  Moreover, sensitivity analysis can be peformed to
determine how choice of $\hat \Delta$ affects inference.  One
limitation of the approach presented here is that only fully
parametric hazard models may be fit to the data. However, a spline
model for the log baseline hazard affords sufficient model
flexibility.

@@

One potential criticism for~\eqref{eq:hazard} is that the health
process may be measured with error.
A common mathematical strategy for joint models is to consider an
unobservable, latent process~$\eta_i$ such that~$\bfT_i \indep x_i
\given \eta_i$. Take~$\eta_i$ to be a zero-mean Gaussian process with
\begin{equation}\label{eq:jm}
x_i(t) = \eta_i (t) + \epsilon_i (t),\quad \text{ and } \log h_i (t
\given \eta ) = \log h_0 (t) + g_t \left( H_{i,t}^N \right)^{\prime}
\alpha + \int_{t-\Delta}^t \eta_i (s) \beta (s) ds
\end{equation}
where~$\epsilon (t)$ is a white-noise measurement error. Thus,
$\eta_i (t)$ is the ``true and unobserved value of the longitudinal
outcome'' \citep[Sec. 2.1, pp.3]{Rizopoulos2010}. The conditional
density function of the event process~$\bfT_i$ given~$\bfx_i$ is
\[
\pr \left ( \bfT_i = \bft \Given \H_{i,\tau_i}^X \right) = E \left(
  \prod_{j=1}^{k_i} h_i (t_j \given \eta) \exp \left(
    \int_{0}^{\tau_i} h_i (s \given \eta ) ds \right) \given \bfx_i \right).
\]
Therefore, the conditional density function depends not only on past
$x$-values, but also on future $x$-values.
That is, the processes~$\bfx$ and~$\bfT$ do not satisfy
\emph{independent evolution}~\citep{DempseyPMCC2}.
This is quite unnatural, as~\eqref{eq:jm} suggests the instantaneous
risk of an event at time~$t$ depends on future values of the sensor
process. To alleviate this issue, we employ the strategy of
taking~$\epsilon (t) \equiv 0$ and treating~$\bfx_i$ as a mean-zero
Gaussian process. The proposed inferential procedure only relies on
the Gaussian assumption when~$\bfx$ is observed with missingness (see
Section~\ref{section:missingdata}).

@@

For the sake of notational simplicity, we leave the dependency of the
conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i
(t; \theta)$. The component of the log-likelihood related to the event
process is then given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i}
    \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_{i} \left( \tau_i; \theta \right) \right)
\]
where~$H_{i} (\tau_i ; \theta) = \int_{0}^{\tau_i} h_{i} (t; \theta)
dt$. Solving the associated score equations~$U_n (\theta) = {\bf 0}$
yields the maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} \frac{h^{(1)}_i
    (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_{i} (\tau_i;
  \theta) \right)
\]
where~$h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (\tau_{i}; \theta)$
are derivatives with respect to $\theta$.

@@

An estimator is design-unbiased if its expectation is equal to that
parameter under the probability distribution induced by the sampling
design~\citep{Cassel1977}. Let~$D_i \subset [0,t_i]$ denote the set of
subsampled points.  Note, by design, this set is distinct from the set
of event times, i.e.,~$\bfT_i \cap D_i = \emptyset$.  Under
subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson
estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in D_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (u) }
\]
An alternative design-unbiased estimator of the derivative of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup D_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested
by~\cite{Waagepetersen2008}.  This estimator depends on the
superposition of the event and subsampling processes.
Proposition~\ref{prop:optimal} shows the estimator for~$\theta$
associated with using~\eqref{eq:WPest} is the most efficient within a
suitable class of estimators for the derivative of the cumulative
hazard function (including the Horvitz-Thompson estimator).
Therefore, we restrict our attention to~\eqref{eq:WPest} for the
remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting estimating equations can be written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( u; \theta)}  - \sum_{u \in D_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built
via plug-in of the design-unbiased estimator of the derivative of the
cumulative hazard given in~\eqref{eq:WPest}.

@@

Probabilistic subsampling converts the single sensor stream~$\bfx_i$
into a sequence of functions observed repeatedly at sampled
times~$D_i$ and event times~$\bfT_i$. Such a data structure is
commonly referred to as \emph{longitudinal functional
  data}~\citep{Xiao2013, GoldSmith2015}.
Given the large increase in longitudinal functional data in recent
years, corresponding analysis has received much recent
attention~\citep{Morris2003, MorrisCarroll2006,
  Baladandayuthapani2008, Di2009, Greven2010, Staicu2010,
  ChenMuller2012, LiGuan2014}.
Here, we combine work by~\cite{Park2018} and~\cite{Goldsmith2011} to
construct a computationally efficient penalized functional method for
solving the estimation equations~$\hat U_n (\theta)$.

@@

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0
\leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup D_i$.
We use the sandwich smoother~\citep{Xiao2013} to estimate the
mean~$\mu(s,t) = \E [ X(t,s)]$.
Alternative bivariate smoothers exist, such as the kernel-based local
linear smoother~\citep{Hastie2009} bivariate tensor product
splines~\citep{Wood2006} and the bivariate penalized spline
smoother~\citep{MarxEilers2005}.
The sandwich smoother was chosen for its computational efficiency and
estimation accuracy.
We then define~$\tilde X(s,t) = X(s,t) - \hat \mu(s,t)$ to be the
mean-zero process at each time~$t \in \bfT_i \cup D_i$.

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma (s, s^\prime) = \int_{0}^\tau c( (s,T), (s^\prime, T) ) g(T) dt.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c((s,t), (s^\prime,t))$ is
the covariance function of the windowed covariate history~$X(t,\cdot)$
and $g(T)$ is the joint sampling density of event and subsampled
timepoints.
Estimation of~$\Sigma$ occurs in two steps.
First, the pooled sample covariance is calculated at a set of grid
points:
\[
\tilde \Sigma (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | N_i
    \cup D_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in N_i
    \cup D_i} \tilde X (s_r,t) \tilde X (s_{r^\prime}, t) \right).
\]
Due to our concern over independent evolution as discussed in
section~\ref{section:memproblems}, we do not assume that each
observation is observed with white noise and therefore the diagonal
elements of~$\hat \Sigma$ are not inflated. Second, the
estimator~$\hat \Sigma$ is further smoothed again using the sandwich
smoother~\citep{Xiao2013}.
Note~\cite{Park2018} smooth the off-diagonal elements, while here we
smooth the entire pooled sample covariance matrix. All negative
eigenvalues are set to zero to ensure positive semi-definiteness.
The result is used as an estimator~$\hat \Sigma$ for the pooled
covariance $\Sigma$.

Next, we take the spectral decomposition of the estimated covariance
function; let~$\{ \hat \psi_k (s),\hat \lambda_k \}_{k \geq 1}$ be the
resulting sequence of eigenfunctions and eigenvalues.
The key benefit of the marginal covariance approach is that it allows
us to compute a single, time-invariant basis expansion; this reduces
the computational burden by avoiding the three dimensional covariance
function (i.e., covariance depends on~$t$) and associated spectral
decomposition in methods considered by \cite{ChenMuller2012}.
Using the Karhunen-Lo{\`e}ve decomposition, we can represent~$X(t,s)$
by
\[
X(s,t) = \hat \mu(s,t) + \sum_{k=1}^{\infty} \hat c_{i,k} (t) \hat
\psi_k (s) \approx \hat \mu(s,t) + {\bf c}_{i} (t)^\top \mathbold{\hat
  \psi} (s)
\]
where $\hat c_{i,k} (t) = \int_{t-\Delta}^t \tilde X_i (s,t) \hat
\psi_k (s) ds$, ${\bf c}_i (t) = (c_{i,1} (t), \ldots, c_{i,K_x}
(t))^\top$, $\mathbold{\hat \psi} (s) = (\hat \psi_1 (s), \ldots, \hat
\psi_{K_x} (s))^\top$, and~$K_x < \infty$ is the truncation level of
the infinite expansion. Following~\cite{Goldsmith2011}, we set~$K_x$
to satisfy identifiability constraints (see Section~\ref{section:beta}
for details).

@@

The next step of our method is modeling~$\beta(s)$. Here, we leverage
ideas from the penalized spline literature~\citep{Ruppert2003,
  Wood2006book}. Let~$\mathbold{\phi} (s) = \{ \phi_1 (s), \ldots,
\phi_{K_b} (x) \}$ be a spline basis and assume that~$\beta(s) =
\sum_{j=1}^{K_b} b_j \phi_{j} (s) = \mathbold{\phi} (t) {\bf b}$
where~${\bf b} = [b_1, \ldots, b_{K_b}]^{\top}$. Thus, the integral
in~\eqref{eq:hazardlinear} can be restated as
\begin{align*}
\int_{t-\Delta}^t X(s,t) \beta(s) ds
  &\approx \int_{t-\Delta}^t \left[ \hat \mu(s,t) + {\bf c} (t)^\top
    \mathbold{\hat \psi} (s) \right] \times \left[
    \mathbold{\phi} (s) {\bf b} \right] \\
  &= [ M_{i,t}^\prime + C_{i,t}^\prime J_{\hat \psi, \phi} ] {\bf b}
\end{align*}
where~$M_{i,t} = (M_{i,1,t}, \ldots, M_{i,K_b,t})$, $M_{i,j,t}
= \int_{t-\Delta}^t \hat \mu (s,t) \phi_j (s)$, and~$J_{\hat \psi,
  \phi}$ is a $K_x \times K_b$ dimensional matrix with the~$(k,l)$th
entry is equal to~$\int_{0}^\Delta \hat \psi_k (s) \phi_l (s)
ds$~\citep{RamsaySilverman2005}.

Given the basis for~$\beta(t)$, the model depends on choice of
both~$K_b$ and~$K_x$.  We follow~\cite{Ruppert2002} by choosing $K_b$
large enough to prevent undersmoothing and~$K_x \geq K_b$ to satisfy
identifability constraints.
While our theoretical analysis considers truncation levels that depend
on~$n$, in practice, we follow the simple rule of thumb and set~$K_b
= K_x = 35$. As long as the choices of~$K_x$ and $K_b$ are large
enough, their impact on estimation is typically negligible.
%We formalize this statement in Theorem~\ref{thm:truncation}.
Below, we will exploit a connection between~\eqref{eq:approxscore}
and score equations for a logistic regression model.  Before moving
on, we introduce notation:
\begin{equation}
\label{eq:approx_hazard}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) \approx
\exp \left( Z_{i,t}^\top \gamma + g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + M_{i,t}^\top {\bf b} + C_{i,t}^\top J_{\hat \psi, \phi} {\bf b} \right)
= \exp \left( W_{i,t}^\top \theta \right),
\end{equation}
where~$\theta = (\gamma, \alpha, {\bf b})$ and~$\exp (
Z_{i,t}^\top \gamma) =: h_0 (t)$ is the parameterized baseline
intensity function. We write~$\tilde U_n (\theta)$ to denote
the approximate score function when substituting
in~\eqref{eq:approx_hazard} for~\eqref{eq:hazardlinear}.

>>>>

\def\S{\mathcal{S}}
\def\indep{\mathrel{\rlap{$\perp$}\kern1.6pt\mathord{\perp}}}
\def\R{\mathcal{R}}
\def\H{\mathcal{H}}
\def\E{\mathbb{E}}
\def\Y{{\bf Y}}
\def\Cov{\text{Cov}}
\def\one{{\bf 1}}
\def\diag{\text{diag}}
\def\given{\, | \,}
\def\Given{\, \big | \,}
\def\Nat{\mathbb{N}}
\def\Real{\mathbb{R}}
\def\bft{{\bf t}}
\def\bfx{{\bf x}}
\def\bfp{{\bf p}}
\def\bfT{{\bf T}}
\def\bfD{{\bf D}}
\def\dotminussym#1#2{%
  \setbox0=\hbox{$\m@th#1-$}%
  \kern.5\wd0%
  \hbox to 0pt{\hss\hbox{$\m@th#1-$}\hss}%
  \raise.6\ht0\hbox to 0pt{\hss$\m@th#1.$\hss}%
  \kern.5\wd0}
\newcommand{\dotminus}{\mathbin{\mathpalette\dotminussym{}}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

Let~$H_{i,t}^{NX} = H_{i,t}^{N} \otimes H_{i,t}^{X}$ be the $\sigma$-field generated by all past values~$(N_i (s), x_i (s))_{0 \leq s \leq t}$. In this paper, the instantaneous risk of an event at time~$t$ is assumed to depend on the health process, time-in-study, and the event history through a fully parametric conditional hazard function:
\begin{equation}
\label{eq:hazard}
h_i \left( t \Given H_{i,t}^{NX} ; \theta \right) =
\lim_{\delta \to 0} \delta^{-1} \pr \left( N_i(t+\delta) - N_i(t) > 0
  \given H_{i,t}^{NX} \right),
\end{equation}
where~$\theta$ is the parameter vector. Equation~\eqref{eq:hazard} states the hazard may depend on past values of the event history and covariates.  For high frequency physiological data, we assume that current risk is log-additive and depends on a linear functional of the health process over some recent window of time and some pre-specified features of the counting process; that is,
\begin{equation}
\label{eq:hazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_i (s) \beta(s) ds  \right)
\end{equation}
where~$h_0(t;\gamma)$ is a parametrized baseline hazard function,~$\Delta$ is an unknown window-length, and $g_t( H_{i,t}^N ) \in \mathbb{R}^p$ is a $p$-length feature vector summarizing the event-history and time-in-study information. The final term~$\int_{t-\Delta}^t x_i(s) \beta(s) ds$ reflects the unknown linear functional form of the impact of the time-varying covariate on current risk.

An alternative to~\eqref{eq:hazardlinear} would be to construct features from the sensor data history~$f_t ( H_{i,t}^{X}) \in \mathbb{R}^q$ and incorporated these features in the place of the
final term. Our current approach builds linear features of~$H_{i,t}^X$ directly from the integrated history, avoiding the feature construction problem -- a highly nontrivial issue for high frequency time-series data.  The main caveat is the additional parameter~$\Delta$; however, as long as the estimated~$\hat \Delta$ exceeds~$\Delta$, then resulting estimation is unbiased albeit at a loss of efficiency.  Moreover, sensitivity analysis can be performed to determine how choice of $\hat \Delta$ affects inference.  One limitation of the approach presented here is that only fully parametric hazard models may be fit to the data. However, a spline model for the log baseline hazard affords sufficient model flexibility.

@@

One potential criticism for~\eqref{eq:hazard} is that the health process may be measured with error. A common mathematical strategy for joint models is to consider an unobservable, latent process~$\eta_i$ such that~$\bfT_i \indep x_i \given \eta_i$, i.e., the two processes are conditionally independent given the latent trajectory. For example, take~$\eta_i$ to be a zero-mean Gaussian process with
\begin{equation}\label{eq:jm}
x_i(t) = \eta_i (t) + \epsilon_i (t),\quad \text{ and } \log h_i (t
\given \eta ) = \log h_0 (t) + g_t \left( H_{i,t}^N \right)^{\prime}
\alpha + \int_{t-\Delta}^t \eta_i (s) \beta (s) ds
\end{equation}
where~$\epsilon_i(t)$ is a white-noise measurement error term. Thus, $\eta_i (t)$ is the ``true and unobserved value of the longitudinal outcome'' \citep[Sec. 2.1, pp.3]{Rizopoulos2010}. The conditional survival function of the $j$th event~$T_j$ given~$\bfx_i$ and all prior events~$\bfT_{-j} := \{ T_1, \ldots, T_{j-1} \}$ is
\[
\pr \left ( T_j > t+s \Given \H_{i,\tau_i}^X, \bfT_{-j} = \bft_{-j}, T_j > t \right) = E \left( - \exp \left( \int_{t}^{t+s} h_i (u \given \eta ) du \right) \Given \bfx_i, \bfT_{-j} = \bft_{-j} \right),
\]
where the expectation is a Gaussian integral, albeit infinite-dimensional.  For most choices of covariance structure, if the white-noise error term is non-zero then the above calculation will show the conditional survival function depends not only on past $x$-values, but also on future $x$-values, i.e.,~$\bfx$ and~$\bfT$ do not satisfy \emph{independent evolution}~\citep{DempseyPMCC2}. This is quite unnatural, as~\eqref{eq:jm} suggests the instantaneous risk of an event at time~$t$ depends on future values of the sensor process. To ensure independent evolution, we employ the strategy of setting~$\epsilon (t) \equiv 0$ and treating~$\bfx_i$ as a mean-zero Gaussian process. The proposed inferential procedure will only rely on the Gaussian assumption when~$\bfx$ is not fully observed (see Section~\ref{section:missingdata}).

@@

For the sake of notational simplicity, we leave the dependency of the conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i (t; \theta)$. The component of the log-likelihood related to the event process is then given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i}
    \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_{i} \left( \tau_i; \theta \right) \right)
\]
where~$H_{i} (\tau_i ; \theta) = \int_{0}^{\tau_i} h_{i} (t; \theta) dt$ is the cumulative hazard function. Solving the associated score equations~$U_n (\theta) = {\bf 0}$ yields the maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} \frac{h^{(1)}_i
    (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_{i} (\tau_i;
  \theta) \right)
\]
where~$h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (\tau_{i}; \theta)$ are derivatives with respect to $\theta$.

@@

An estimator is design-unbiased if its expectation is equal to that parameter under the probability distribution induced by the sampling design~\citep{Cassel1977}. Let~$\bfD_i \subset [0,t_i]$ denote the set of subsampled points.  Note, by construction, this set is distinct from the set of event times, i.e.,~$\bfT_i \cap \bfD_i = \emptyset$.  Under subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in \bfD_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (u) }
\]
An alternative design-unbiased estimator of the derivative of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup \bfD_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested by~\cite{Waagepetersen2008}.  This estimator depends on the superposition of the event and subsampling processes. Proposition~\ref{prop:optimal} shows the estimator for~$\theta$ associated with using~\eqref{eq:WPest} is the most efficient within a suitable class of estimators for the derivative of the cumulative hazard function (including the Horvitz-Thompson estimator). Therefore, we restrict our attention to~\eqref{eq:WPest} for the remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting estimating equations can be written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( u; \theta)}  - \sum_{u \in D_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built via plug-in of the design-unbiased estimator of the derivative of the cumulative hazard given in~\eqref{eq:WPest}.

\begin{rmk}[Connection to design-based estimation]
The approximate score equations given by~\eqref{eq:approxscore} arise in design-based inference of point processes.  Design-based inference is common for spatial point processes~\citep{Waagepetersen2008} where the spatial varying covariate is observed at a random sample of locations. It is common in mobile health where ecological momentary assessments~\citep{Rathbun2012,Rathbun2016} are used to randomly sample individuals at various time-points to assess their emotional state.  In the current setting, we leverage these ideas to form a subsampling protocol that can substantially reduce computationally complexity.  Therefore, the purpose is quite different. Moreover, the dependence of the intensity function on the recent history of sensor values leads to additional complications that must be addressed.
\end{rmk}

@@

Probabilistic subsampling converts the single sensor stream~$\bfx_i$ into a sequence of functions observed repeatedly at sampled times~$\bfD_i$ and event times~$\bfT_i$. Such a data structure is commonly referred to as \emph{longitudinal functional data}~\citep{Xiao2013, GoldSmith2015}.
Given the large increase in longitudinal functional data in recent years, corresponding analysis has received much recent attention~\citep{Morris2003, MorrisCarroll2006, Baladandayuthapani2008, Di2009, Greven2010, Staicu2010, ChenMuller2012, LiGuan2014}. Here, we combine work by~\cite{Park2018} and~\cite{Goldsmith2011} to construct a computationally efficient penalized functional method for solving the estimation equations~$\hat U_n (\theta)$.

@@

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. We use the sandwich smoother~\citep{Xiao2013} to estimate the mean~$\mu_y(s,t) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009} bivariate tensor product splines~\citep{Wood2006} and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(s,t) = X(s,t) - \hat \mu_y(s,t)$ to be the mean-zero process at each time~$t \in \bfT_i \cup D_i$.

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma_y (s, s^\prime) = \int_{0}^\tau c_y( (s,T), (s^\prime, T) ) g_y(T) dt.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c_y((s,t), (s^\prime,t))$ is the covariance function of the windowed covariate history~$X(t,\cdot)$ and $g_y(T)$ is the intensity function for event ($y=1$) and subsampled ($y=0$) times respectively. Estimation of~$\Sigma_y$ occurs in two steps. For simplicity, we present the steps for subsampled times (i.e., $y=0$) but the steps are the same for event times as well. First, the pooled sample covariance is calculated at a set of grid points:
\[
\tilde \Sigma_0 (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | \bfD_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in \bfD_i} \tilde X (s_r,t) \tilde X (s_{r^\prime}, t) \right).
\]
Due to our concern over independent evolution as discussed in section~\ref{section:memproblems}, we do not assume that each observation is observed with white noise and therefore the diagonal
elements of~$\hat \Sigma_0$ are not inflated. Second, the estimator~$\hat \Sigma$ is further smoothed again using the sandwich smoother~\citep{Xiao2013}. Note~\cite{Park2018} smooth the off-diagonal elements, while here we smooth the entire pooled sample covariance matrix. All negative eigenvalues are set to zero to ensure positive semi-definiteness. The result is used as an estimator~$\hat \Sigma_0$ for the pooled covariance $\Sigma_0$.

Next, we take the spectral decomposition of the estimated covariance function; let $\{ \hat \psi^{(0)}_k (s),\hat \lambda^{(0)}_k \}_{k \geq 1}$ be the resulting sequence of eigenfunctions and eigenvalues. The key benefit of the marginal covariance approach is that it allows us to compute a single, time-invariant basis expansion; this reduces the computational burden by avoiding the three dimensional covariance function (i.e., covariance depends on~$t$) and associated spectral decomposition in methods considered by \cite{ChenMuller2012}. Using the Karhunen-Lo{\`e}ve decomposition, we can represent~$X(t,s)$ for $t \in \bfT_i \cap \bfD_i$ by
\[
X(s,t) = \hat \mu_y (s,t) + \sum_{k=1}^{\infty} \hat c^{(y)}_{i,k} (t) \hat
\psi^{(y)}_k (s) \approx \hat \mu_y (s,t) + {\bf c}^{(y)}_{i} (t)^\top \mathbold{\hat
  \psi}^{(y)} (s)
\]
where $\hat c^{(y)}_{i,k} (t) = \int_{t-\Delta}^t \tilde X_i (s,t) \hat \psi^{(y)}_k (s) ds$, ${\bf c}^{(y)}_i (t) = (c^{(y)}_{i,1} (t), \ldots, c^{(y)}_{i,K_x} (t))^\top$, $\mathbold{\hat \psi}^{(y)} (s) = (\hat \psi^{(y)}_1 (s), \ldots, \hat \psi^{(y)}_{K_x} (s))^\top$, and~$K_x < \infty$ is the truncation level of the infinite expansion. Following~\cite{Goldsmith2011}, we set~$K_x$ to satisfy identifiability constraints (see Section~\ref{section:beta} for details). In subsequent sections, we leave the dependence on $y$ (i.e., whether $t \in \bfT_i$ or $\in \bfD_i$) implicit unless required for notational simplicity.

@@

The next step of our method is modeling~$\beta(s)$. Here, we leverage ideas from the penalized spline literature~\citep{Ruppert2003, Wood2006book}. Let~$\mathbold{\phi} (s) = \{ \phi_1 (s), \ldots, \phi_{K_b} (x) \}$ be a spline basis and assume that~$\beta(s) = \sum_{j=1}^{K_b} b_j \phi_{j} (s) = \mathbold{\phi} (t) {\bf b}$ where~${\bf b} = [b_1, \ldots, b_{K_b}]^{\top}$. Thus, the integral in~\eqref{eq:hazardlinear} can be restated as
\begin{align*}
\int_{t-\Delta}^t X(s,t) \beta(s) ds
  &\approx \int_{t-\Delta}^t \left[ \hat \mu(s,t) + {\bf c} (t)^\top
    \mathbold{\hat \psi} (s) \right] \times \left[
    \mathbold{\phi} (s) {\bf b} \right] \\
  &= [ ( M_{i,t} )^\top + (C_{i,t})^\top J_{\hat \psi, \phi} ] {\bf b}
\end{align*}
where~$M_{i,t} = (M_{i,1,t}, \ldots, M_{i,K_b,t})$, $M_{i,j,t} = \int_{t-\Delta}^t \hat \mu (s,t) \phi_j (s)$, and~$J_{\hat \psi, \phi}$ is a $K_x \times K_b$ dimensional matrix with the~$(k,l)$th entry is equal to~$\int_{0}^\Delta \hat \psi_k (s) \phi_l (s) ds$~\citep{RamsaySilverman2005}.

Given the basis for~$\beta(t)$, the model depends on choice of both~$K_b$ and~$K_x$.  We follow~\cite{Ruppert2002} by choosing $K_b$ large enough to prevent under-smoothing and~$K_x \geq K_b$ to satisfy identifiability constraints. While our theoretical analysis considers truncation levels that depend on~$n$, in practice, we follow the simple rule of thumb and set~$K_b = K_x = 35$. As long as the choices of~$K_x$ and $K_b$ are large enough, their impact on estimation is typically negligible. Below, we will exploit a connection between~\eqref{eq:approxscore} and score equations for a logistic regression model.  Before moving on, we introduce some additional notation.  First, let the baseline hazard $h_0(t; \gamma) := \exp(Z_{i,t}^\top \gamma)$ where~$Z_{i,t} \in \mathbb{R}^{q}$ parametrizes some piecewise-constant baseline   be :
\begin{equation}
\label{eq:approx_hazard}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) \approx
\exp \left( Z_{i,t}^\top \gamma + g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + M_{i,t}^\top {\bf b} + C_{i,t}^\top J_{\hat \psi, \phi} {\bf b} \right)
= \exp \left( W_{i,t}^\top \theta \right),
\end{equation}
where~$\theta = (\gamma, \alpha, {\bf b})$ and~$\exp ( Z_{i,t}^\top \gamma) =: h_0 (t)$ is the parameterized baseline intensity function. We write~$\tilde U_n (\theta)$ to denote the approximate score function when substituting in~\eqref{eq:approx_hazard} for~\eqref{eq:hazardlinear}.
